---
title: "Rise and Fall of Scottish COVID-19: Sequence Data Report"
author: "Samantha Lycett"
date: "27/05/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
library(knitr)
#library(kableExtra)
library(tibble)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=6)
library(ape)
```

```{r, include=FALSE, echo=FALSE}
# load some custom functions
#Rpath <- "~/Documents//GitLab//RiseFall-Scottish-COVID19//coronavirus_2020_r//"
Rpath <- ""
source(paste(Rpath,"getEl.R",sep=""))
source(paste(Rpath,"calcDecimalDate.R",sep=""))
source(paste(Rpath,"get_BEAST_cols.R",sep=""))
source(paste(Rpath,"get_udates.R",sep=""))
source(paste(Rpath,"birdSpecies_and_continents.R",sep=""))
```

```{r, include=FALSE, echo=FALSE}
## THIS CAN BE CHANGED - dataDate is the date of the dataset, and is in the path and file names
dataDate <- "2020-05-22"
```

```{r, include=FALSE, echo=FALSE}
# set up paths
rootPath <- "~/Documents//data//Coronavirus_2020/COG_downloads//"
path2    <- "microreact"
ext      <-"_metadata_private.csv"
metaPath <- paste(rootPath,dataDate,"//",path2,"//",sep="")
metaName <- paste("cog_global_",dataDate,ext,sep="")
treeName <- paste("cog_global_",dataDate,"_tree.newick",sep="")
```

```{r, include=FALSE, echo=FALSE}
# read metadata
metadata <- read.csv(paste(metaPath,metaName,sep=""))
nseqs <- length(metadata[,1])
```

```{r, include=FALSE, echo=FALSE}
# configure the metadata information

# basic continents
region<- unGeoScheme(metadata$country)
continent <- region
# re-configure so is like the GISAID
continent[grep("Oceania", continent)] <- "Oceania"
continent[grep("Asia", continent)]    <- "Asia"
continent[grep("Europe", continent)]  <- "Europe"
continent[grep("Africa", continent)]  <- "Africa"
continent[grep("NorthernAmerica", continent)] <- "North America"
continent[grep("SouthernAmerica", continent)] <- "South America"
continent[grep("CentralAmerica", continent)] <- "Central America"
ccols <- colourize_BEAST_cols(continent)
cont_tbl <- table(continent)
ctxt  <- paste(rownames(cont_tbl)," (",cont_tbl,")",sep="")

# country_state and UK
country       <- metadata$country
country_state <- paste(metadata$adm1)
u_uk_states   <- c("England","Wales","Scotland","Northern Ireland")
country_state[grep("UK-ENG", country_state)] <- "England"
country_state[grep("UK-SCT", country_state)] <- "Scotland"
country_state[grep("UK-WLS", country_state)] <- "Wales"
country_state[grep("UK-NIR", country_state)] <- "Northern Ireland"
country_state[which(country_state=="")] <- "-"
country_state <- factor(country_state, levels=c(u_uk_states,"-"))
uk_cols     <- colourize_BEAST_cols(country_state, ustates=u_uk_states, 
                                    ucols=c(get_BEAST_cols(4,bright=0.8, sat=0.7),"grey90"))
uk_inds     <- which(country=="UK")
uk_tbl <- table(country_state[uk_inds])
uk_txt <- paste(rownames(uk_tbl)," (",uk_tbl,")",sep="")
uk_txt <- uk_txt[1:4]

# uk_county
uk_county <- paste(metadata$adm2)
uk_county[grep("UNKNOWN",uk_county)] <- "UNKNOWN"
uk_county[which(uk_county=="")] <- "UNKNOWN"
u_uk_county <- setdiff(sort(unique(uk_county)),"UNKNOWN")
uk_county <- factor(uk_county, levels=c(u_uk_county,"UNKNOWN"))

# scotland
scot_inds <- which(country_state=="Scotland")
u_scot    <- c(setdiff(sort(unique(uk_county[scot_inds])),"UNKNOWN"),"UNKNOWN")
scot_cols <- colourize_BEAST_cols(uk_county, ustates=u_scot, 
                                  ucols=c(get_BEAST_cols(length(u_scot)-1,bright=0.8, sat=0.7),"grey90"))
scot_tbl  <- table(factor(paste(uk_county[scot_inds]), levels=u_scot))
scot_txt  <- paste(rownames(scot_tbl)," (",scot_tbl,")",sep="")

# decimal dates
# correction
metadata$sample_date[which(metadata$sample_date=="2022-04-05")] <- "2020-04-05"
decDates <- apply(as.matrix(metadata$sample_date), 1, calcDecimalDate_fromTxt, sep="-", dayFirst=FALSE)

maxDate   <- max(decDates)
max_ii    <- which.max(decDates)
min_ii    <- which.min(decDates)
min_date  <- as.integer(strsplit(paste(metadata$sample_date[min_ii]),"-")[[1]])
if (min(decDates) > 2020) {
  min_date<- c(2020,1,1)
}
  
max_date  <- as.integer(strsplit(paste(metadata$sample_date[max_ii]),"-")[[1]])
dates_res <- get_udates(min_date=min_date,max_date=max_date)
```

```{r, include=FALSE, echo=FALSE}
# define custom plot functions
do_continent_pie <- function(cont_tbl=cont_tbl, ccols=ccols, 
                             ctxt=ctxt, dataDate=dataDate, show.legend=TRUE) {
  op <- par(mar=c(0,0,2,0))
  pie(cont_tbl, col=ccols$ucols)
  title(paste("Number of sequences",dataDate))
  if (show.legend) {
    if (length(ctxt)<10) {
      legend("topleft",ctxt,pch=22,bty="n",pt.bg=ccols$ucols)
    } else {
      nn <- length(ctxt)
      nn2<- ceiling(nn/2)
      legend("topleft",ctxt[1:nn2],pch=22,bty="n",pt.bg=ccols$ucols[1:nn2])
      legend("topright",ctxt[(nn2+1):nn],pch=22,bty="n",pt.bg=ccols$ucols[(nn2+1):nn])
    }
  }
  par(op)
}

do_continent_bars <- function(continent, ctxt=ctxt, ccols=ccols, dates=dates, 
                              dataDate=dataDate, dates_res=dates_res) {
  udate    <- dates_res$udate
  seldates <- dates_res$seldates
  seldates2<- dates_res$seldates2
  xlabs    <- dates_res$xlabs

  barplot(  table( continent, factor(dates,levels=udate ) ), 
            names=xlabs, col=ccols$ucols, xlab="Sample date", ylab="Number of sequences" )
  legend("topleft",ctxt,pch=22,bty="n",pt.bg=ccols$ucols)
  title(paste("Number of sequences",dataDate))
}

do_uk_bars <- function(country_state, uk_inds=uk_inds, uk_txt=uk_txt, uk_cols=uk_cols, dates=dates, 
                              dataDate=dataDate, dates_res=dates_res, show.legend=TRUE) {
  udate    <- dates_res$udate
  seldates <- dates_res$seldates
  seldates2<- dates_res$seldates2
  xlabs    <- dates_res$xlabs

  barplot(  table( country_state[uk_inds], factor(dates[uk_inds],levels=udate ) ), 
            names=xlabs, col=uk_cols$ucols, xlab="Sample date", ylab="Number of sequences" )
  
  if (show.legend) {
    if (length(uk_txt)<10) {
      legend("topleft",uk_txt,pch=22,bty="n",pt.bg=uk_cols$ucols)
    } else {
      nn <- length(uk_txt)
      nn2<- ceiling(nn/2)
      legend("topleft",uk_txt[1:nn2],pch=22,bty="n",pt.bg=uk_cols$ucols)
      legend("topright",uk_txt[(nn2+1):nn],pch=22,bty="n",pt.bg=uk_cols$ucols)
    }
  }
  title(paste("Number of sequences",dataDate))
}
```

```{r, echo=FALSE, include=FALSE}
# read global newick tree
tr <- read.tree( paste(metaPath,treeName,sep=""))
```


# Project Details
### Title: Tracking the rise and fall of Scottish SARS-CoV-2 / COVID-19 using virus sequences
### CSO project number: COV/EDI/20/11

# Sequence Data Report
This report summarises the data set of `r dataDate` made available on the MRC CLIMB server via the COVID-19 Genomics UK Consortium (COG) ( https://www.cogconsortium.uk )

The data set comprises whole genome sequences from the England, Scotland, Wales and Northern Ireland sequenced by several centres in the consortium; and GISAID ( https://www.gisaid.org ) sequences from other countries included as background information.  The UK sequences are also being submitted to the GISAID database as part of the sequencing pipeline (but they are not included twice in this file).

# Global Sequences
There are `r nseqs` sequences in the `r dataDate` data set.

```{r, echo=FALSE}
do_continent_pie(cont_tbl=cont_tbl, ccols=ccols, ctxt=ctxt, dataDate=dataDate)
```

```{r, echo=FALSE}
do_continent_bars(continent=continent, ctxt=ctxt, ccols=ccols, 
                  dates=metadata$sample_date, dataDate=dataDate,
                  dates_res=dates_res)
```

# UK Sequences
There are `r length(uk_inds)` UK sequences in the `r dataDate` data set, representing `r format(100*length(uk_inds)/nseqs, digits=2)`% of the total global sequences.

```{r, echo=FALSE}
do_uk_bars(country_state, uk_inds=uk_inds, uk_txt=uk_txt,
           uk_cols=uk_cols, dates=metadata$sample_date, 
           dataDate=dataDate, dates_res=dates_res)
```

\newpage

```{r, echo=FALSE}
kable(table(uk_county[uk_inds],factor(country_state[uk_inds],levels=u_uk_states)), 
      caption="UK Sequences by Region" ) #%>% kable_styling(bootstrap_options = "striped")
```

\newpage

# Scottish Sequences
There are `r length(scot_inds)` Scottish sequences in the `r dataDate` data set, representing `r format(100*length(scot_inds)/length(uk_inds), digits=2)`% of the UK sequences from COG consortium.

```{r, echo=FALSE}
do_continent_pie(cont_tbl=scot_tbl, ccols=scot_cols, ctxt=scot_txt, dataDate=dataDate)
```

```{r, echo=FALSE}
do_uk_bars(uk_county, uk_inds=scot_inds, uk_txt=scot_txt,
           uk_cols=scot_cols, dates=metadata$sample_date, 
           dataDate=dataDate, dates_res=dates_res, show.legend=FALSE)
```

\newpage

```{r, echo=FALSE}
kable( table(  factor(metadata$adm2[scot_inds], levels=u_scot),metadata$epi_week[scot_inds]), caption="Scottish sequences by epi-week"  ) #%>% kable_styling(bootstrap_options = "striped")
```

\newpage

```{r, echo=FALSE}
kable( table(  factor(metadata$adm2[scot_inds], levels=u_scot),paste(metadata$lineage[scot_inds])), caption="Scottish sequences by lineage", font_size=8  ) #%>% kable_styling(bootstrap_options = "striped", font_size = 8)
```


