---
title: "SARS-CoV-2 Imports and Exports"
subtitle: "Scotland and India"
author: "Samantha Lycett"
date: "29/05/2020"
output: html_document
---

```{r setup, include=FALSE}
library(knitr)
#library(kableExtra)
library(tibble)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width=12, fig.height=6)
library(ape)
```

```{r, include=FALSE, echo=FALSE}
# load some custom functions
#Rpath <- "~/Documents//GitLab//RiseFall-Scottish-COVID19//coronavirus_2020_r//"
Rpath <- ""
source(paste(Rpath,"getEl.R",sep=""))
source(paste(Rpath,"calcDecimalDate.R",sep=""))
source(paste(Rpath,"get_BEAST_cols.R",sep=""))
source(paste(Rpath,"get_udates.R",sep=""))
source(paste(Rpath,"birdSpecies_and_continents.R",sep=""))
```

```{r, include=FALSE, echo=FALSE}
## THIS CAN BE CHANGED - dataDate is the date of the dataset, and is in the path and file names
dataDate <- "2020-05-22"
```

```{r, include=FALSE, echo=FALSE}
# set up paths
# data is from the public microreact site
rootPath <- "~/Documents//data//Coronavirus_2020//Microreact//"
n2       <- "_microreact-project-cogconsortium-"
ext      <-"data.csv"
metaPath <- paste(rootPath,dataDate,"//",sep="")
metaName <- paste(dataDate,n2,ext,sep="")
treeName <- paste(dataDate,n2,"tree.nwk",sep="")
```

```{r, include=FALSE, echo=FALSE}
# read metadata
# only country and adm1
metadata <- read.csv(paste(metaPath,metaName,sep=""))
nseqs <- length(metadata[,1])
```

```{r, include=FALSE, echo=FALSE}
# configure the metadata information

# basic continents
region<- unGeoScheme(metadata$country)
continent <- region
# re-configure so is like the GISAID
continent[grep("Oceania", continent)] <- "Oceania"
continent[grep("Asia", continent)]    <- "Asia"
continent[grep("Europe", continent)]  <- "Europe"
continent[grep("Africa", continent)]  <- "Africa"
continent[grep("NorthernAmerica", continent)] <- "North America"
continent[grep("SouthernAmerica", continent)] <- "South America"
continent[grep("CentralAmerica", continent)] <- "Central America"
ccols <- colourize_BEAST_cols(continent)
cont_tbl <- table(continent)
ctxt  <- paste(rownames(cont_tbl)," (",cont_tbl,")",sep="")

# country_state and UK
country       <- metadata$country
country_state <- paste(metadata$adm1)
u_uk_states   <- c("England","Wales","Scotland","Northern Ireland")
country_state[grep("UK-ENG", country_state)] <- "England"
country_state[grep("UK-SCT", country_state)] <- "Scotland"
country_state[grep("UK-WLS", country_state)] <- "Wales"
country_state[grep("UK-NIR", country_state)] <- "Northern Ireland"
country_state[which(country_state=="")] <- "-"
country_state <- factor(country_state, levels=c(u_uk_states,"-"))
uk_cols     <- colourize_BEAST_cols(country_state, ustates=u_uk_states, 
                                    ucols=c(get_BEAST_cols(4,bright=0.8, sat=0.7),"grey90"))
uk_inds     <- which(country=="UK")
uk_tbl <- table(country_state[uk_inds])
uk_txt <- paste(rownames(uk_tbl)," (",uk_tbl,")",sep="")
uk_txt <- uk_txt[1:4]

# scotland
scot_inds <- which(country_state=="Scotland")

# decimal dates
# correction
metadata$sample_date[which(metadata$sample_date=="2022-04-05")] <- "2020-04-05"
decDates <- apply(as.matrix(metadata$sample_date), 1, calcDecimalDate_fromTxt, sep="-", dayFirst=FALSE)

maxDate   <- max(decDates)
max_ii    <- which.max(decDates)
min_ii    <- which.min(decDates)
min_date  <- as.integer(strsplit(paste(metadata$sample_date[min_ii]),"-")[[1]])
if (min(decDates) > 2020) {
  min_date<- c(2020,1,1)
}
  
max_date  <- as.integer(strsplit(paste(metadata$sample_date[max_ii]),"-")[[1]])
dates_res <- get_udates(min_date=min_date,max_date=max_date)
```

```{r, include=FALSE, echo=FALSE}
# define custom plot functions
do_continent_pie <- function(cont_tbl=cont_tbl, ccols=ccols, 
                             ctxt=ctxt, dataDate=dataDate, show.legend=TRUE) {
  op <- par(mar=c(0,0,2,0))
  pie(cont_tbl, col=ccols$ucols)
  title(paste("Number of sequences",dataDate))
  if (show.legend) {
    if (length(ctxt)<10) {
      legend("topleft",ctxt,pch=22,bty="n",pt.bg=ccols$ucols)
    } else {
      nn <- length(ctxt)
      nn2<- ceiling(nn/2)
      legend("topleft",ctxt[1:nn2],pch=22,bty="n",pt.bg=ccols$ucols[1:nn2])
      legend("topright",ctxt[(nn2+1):nn],pch=22,bty="n",pt.bg=ccols$ucols[(nn2+1):nn])
    }
  }
  par(op)
}

do_continent_bars <- function(continent, ctxt=ctxt, ccols=ccols, dates=dates, 
                              dataDate=dataDate, dates_res=dates_res) {
  udate    <- dates_res$udate
  seldates <- dates_res$seldates
  seldates2<- dates_res$seldates2
  xlabs    <- dates_res$xlabs

  barplot(  table( continent, factor(dates,levels=udate ) ), 
            names=xlabs, col=ccols$ucols, xlab="Sample date", ylab="Number of sequences" )
  legend("topleft",ctxt,pch=22,bty="n",pt.bg=ccols$ucols)
  title(paste("Number of sequences",dataDate))
}

do_uk_bars <- function(country_state, uk_inds=uk_inds, uk_txt=uk_txt, uk_cols=uk_cols, dates=dates, 
                              dataDate=dataDate, dates_res=dates_res, show.legend=TRUE) {
  udate    <- dates_res$udate
  seldates <- dates_res$seldates
  seldates2<- dates_res$seldates2
  xlabs    <- dates_res$xlabs

  barplot(  table( country_state[uk_inds], factor(dates[uk_inds],levels=udate ) ), 
            names=xlabs, col=uk_cols$ucols, xlab="Sample date", ylab="Number of sequences" )
  
  if (show.legend) {
    if (length(uk_txt)<10) {
      legend("topleft",uk_txt,pch=22,bty="n",pt.bg=uk_cols$ucols)
    } else {
      nn <- length(uk_txt)
      nn2<- ceiling(nn/2)
      legend("topleft",uk_txt[1:nn2],pch=22,bty="n",pt.bg=uk_cols$ucols)
      legend("topright",uk_txt[(nn2+1):nn],pch=22,bty="n",pt.bg=uk_cols$ucols)
    }
  }
  title(paste("Number of sequences",dataDate))
}
```

```{r, echo=FALSE, include=FALSE}
# read global newick tree
tr <- read.tree( paste(metaPath,treeName,sep=""))
```

# Project Details
### Title: Tracking the rise and fall of Scottish SARS-CoV-2 / COVID-19 using virus sequences
### CSO project number: COV/EDI/20/11

# Sequence Data Report
This report summarises the public data set of `r dataDate` made available on Mircoreact (https://microreact.org/project/cogconsortium/) which is part of the MRC CLIMB server, and has data collected and processed via the COVID-19 Genomics UK Consortium (COG) ( https://www.cogconsortium.uk )

The data set comprises a report on whole genome sequences from the England, Scotland, Wales and Northern Ireland sequenced by several centres in the consortium; and GISAID ( https://www.gisaid.org ) sequences from other countries included as background information.  The UK sequences are also being submitted to the GISAID database as part of the sequencing pipeline (but they are not included twice in this file).

# Global Sequences
There are `r nseqs` sequences in the `r dataDate` data set.

```{r, echo=FALSE}
do_continent_pie(cont_tbl=cont_tbl, ccols=ccols, ctxt=ctxt, dataDate=dataDate)
```

```{r, echo=FALSE}
do_continent_bars(continent=continent, ctxt=ctxt, ccols=ccols, 
                  dates=metadata$sample_date, dataDate=dataDate,
                  dates_res=dates_res)
```

# UK and Scottish sequences
There are `r length(uk_inds)` UK sequences in the `r dataDate` data set, representing `r format(100*length(uk_inds)/nseqs, digits=2)`% of the total global sequences.

There are `r length(scot_inds)` Scottish sequences in the `r dataDate` data set, representing `r format(100*length(scot_inds)/length(uk_inds), digits=2)`% of the UK sequences from COG consortium, and `r format(100*length(scot_inds)/nseqs, digits=2)`% of the total global sequences.

```{r, echo=FALSE}
do_uk_bars(country_state, uk_inds=uk_inds, uk_txt=uk_txt,
           uk_cols=uk_cols, dates=metadata$sample_date, 
           dataDate=dataDate, dates_res=dates_res)
```

```{r, echo=FALSE}
do_continent_pie(cont_tbl=uk_tbl, ccols=uk_cols, ctxt=uk_txt, dataDate=dataDate)
```

# Asia and India sequences
```{r, echo=FALSE}
asia_inds <- which(continent=="Asia")
india_inds<- which(metadata$country=="India")
```
There are `r length(asia_inds)` Asia sequences in the `r dataDate` data set, representing `r format(100*length(asia_inds)/nseqs, digits=2)`% of the total global sequences.

Of these, there are `r length(india_inds)` India sequences in the `r dataDate` data set, representing `r format(100*length(india_inds)/length(asia_inds), digits=2)`% of the Asian sequences, and `r format(100*length(india_inds)/nseqs, digits=2)`% of the total global sequences.

```{r, echo=FALSE}
asia_country <- metadata$country[asia_inds]
atbl <- table(paste(asia_country))
atxt <- paste(rownames(atbl)," (",atbl,")",sep="")
acols<- colourize_BEAST_cols(asia_country)
do_continent_bars(continent=asia_country, ctxt=atxt, ccols=acols, 
                  dates=metadata$sample_date[asia_inds], dataDate=dataDate,
                  dates_res=dates_res)
```
```{r, echo=FALSE}
do_continent_pie(cont_tbl=atbl, ccols=acols, ctxt=atxt, dataDate=dataDate)
```

# Scotland and India Compared

Time line for Scotland and India sequences only
```{r, echo=FALSE}
scot_india_country <- c(paste(metadata$country[india_inds]),paste(country_state[scot_inds]))
sitbl <- table(paste(scot_india_country))
sitxt <- paste(rownames(sitbl)," (",sitbl,")",sep="")
sicols<- colourize_BEAST_cols(scot_india_country, sat=0.5, bright=0.9)
do_continent_bars(continent=scot_india_country, ctxt=sitxt, ccols=sicols, 
                  dates=metadata$sample_date[c(india_inds,scot_inds)], dataDate=dataDate,
                  dates_res=dates_res)
```

Tree with Scotland and India sequences marked on
```{r, echo=FALSE}
tinds        <- match(tr$tip.label, metadata$id)
#all( tr$tip.label == metadata$id[tinds])
tr           <- ladderize(tr)
tr$continent <- continent[tinds]
tr$country   <- metadata$country[tinds]
tr$country_state <- country_state[tinds]
tr$lineage  <- metadata$lineage[tinds]
```
```{r, echo=FALSE}
op <- par(mfrow=c(1,2))

ppch <- array(NA,length(tr$tip.label))
#ppch[which(tr$country_state=="Scotland")] <- 21
ppch[which(tr$country=="India")] <- 23
tipcols <- array(NA,length(tr$tip.label))
#tipcols[which(tr$country_state=="Scotland")] <- sicols$ucols[2]
tipcols[which(tr$country=="India")] <- sicols$ucols[1]

plot(tr, show.tip=FALSE, edge.col="grey70")
tiplabels(pch=ppch,bg=tipcols, cex=1.5)
add.scale.bar(2e-4,0,1e-4)
title("Tree with India")

ppch <- array(NA,length(tr$tip.label))
ppch[which(tr$country_state=="Scotland")] <- 21
#ppch[which(tr$country=="India")] <- 23
tipcols <- array(NA,length(tr$tip.label))
tipcols[which(tr$country_state=="Scotland")] <- sicols$ucols[2]
#tipcols[which(tr$country=="India")] <- sicols$ucols[1]

plot(tr, show.tip=FALSE, edge.col="grey70")
tiplabels(pch=ppch,bg=tipcols, cex=1.5)
add.scale.bar(2e-4,0,1e-4)
title("Tree with Scotland")

par(op)

```

Sub trees with Scotland and India per named lineage.

```{r, echo=FALSE}
ulin <- sort(unique(metadata$lineage[india_inds]))
ulin <- setdiff(ulin,c("A","B"))
ulin <- sort(c(ulin,"A.5"))
sts       <- vector("list",length(ulin))
sub_stats <- matrix(0, length(ulin), 6)
rownames(sub_stats) <- ulin
colnames(sub_stats) <- c("Num_Seqs","Num_India","Num_Scotland","Num_Lin_Seqs","Num_Lin_India","Num_Lin_Scotland")
op <- par(mfrow=c(1,2))
for (i in 1:length(ulin)) {
  tips<- which(tr$lineage==ulin[i] & tr$country=="India")
  if (length(tips)==0) {
    tips<- which(tr$lineage==ulin[i] & tr$country_state=="Scotland")
  }
  if (length(tips)>1) {
    anc <- getMRCA(tr,tips)
  } else {
    anc <- tr$edge[which(tr$edge[,2]==tips),1]
  }
  anc      <- tr$edge[which(tr$edge[,2]==anc),1]
  sub_tree <- extract.clade(tr, anc)
  sinds <- match(sub_tree$tip.label, metadata$id)
  sub_tree$country       <- metadata$country[sinds]
  sub_tree$country_state <- country_state[sinds]
  sub_tree$continent     <- continent[sinds]
  sub_tree$lineage       <- metadata$lineage[sinds]
  sub_tree$tipcols       <- array(NA, length(sub_tree$tip.label))
  sub_tree$tipcols[which(sub_tree$country=="India")]          <- sicols$ucols[which(sicols$ustates=="India")]
  sub_tree$tipcols[which(sub_tree$country_state=="Scotland")] <- sicols$ucols[which(sicols$ustates=="Scotland")]
  sub_tree$tipcols2       <- array(hsv(0,0,0.7,0.5), length(sub_tree$tip.label))
  sub_tree$tipcols2[which(sub_tree$country=="India")]          <- "black"
  sub_tree$tipcols2[which(sub_tree$country_state=="Scotland")] <- "black"
  sub_tree$ppch          <- array(24, length(sub_tree$tip.label))
  sub_tree$ppch[which(sub_tree$country=="India")]          <- 23
  sub_tree$ppch[which(sub_tree$country_state=="Scotland")] <- 21
  sub_tree$cex <- array(0.75,length(sub_tree$tip.label))
  sub_tree$cex[which(sub_tree$lineage==ulin[i])] <- 1.5
  sub_stats[i,1] <- length(sub_tree$tip.label)
  sub_stats[i,2] <- length(which(sub_tree$country=="India"))
  sub_stats[i,3] <- length(which(sub_tree$country_state=="Scotland"))
  sub_stats[i,4] <- length(which(sub_tree$lineage==ulin[i]))
  sub_stats[i,5] <- length(which(sub_tree$country=="India" & sub_tree$lineage==ulin[i]))
  sub_stats[i,6] <- length(which(sub_tree$country_state=="Scotland" & sub_tree$lineage==ulin[i]))
  sts[[i]]       <- sub_tree
  
  stxt <- c(paste(ulin[i]," India (",length(which(sub_tree$country=="India" & sub_tree$lineage==ulin[i])),")",sep=""),
            paste(ulin[i]," Scotland (",length(which(sub_tree$country_state=="Scotland" & sub_tree$lineage==ulin[i])),")",sep=""))
  
  if (sub_stats[i,1] < 4000) {
    plot(sub_tree, show.tip=FALSE, edge.col=hsv(0,0,0.7,0.5))
    tiplabels(pch=sub_tree$ppch,bg=sub_tree$tipcols, col=sub_tree$tipcols2, cex=sub_tree$cex)
    add.scale.bar(2e-4,0,1e-4)
    legend("bottomleft",stxt,pch=c(23,21),pt.bg=sicols$ucols,bty="n")
    title(paste(ulin[i],"sub tree with India and Scotland"))
  }
}
par(op)
```

Using a discrete traits model on these subtrees, it is possible to estimate the numbers of imports and exports to Scotland and India.  Here the simplest equal rates discrete trait model is used (most similar to parsimony).  Estimates are dependent upon the number of sequences in each category, and the accuracy of the subtree.

```{r, echo=FALSE}
etol <- 1e-12
utipstate <- c(ccols$ustates,"China","Singapore","India","Italy","Spain",u_uk_states)
focus_states <- c("Scotland","India")
import_tbl_1 <- matrix(-1, length(ulin), length(utipstate))
import_tbl_2 <- matrix(-1, length(ulin), length(utipstate))
export_tbl_1 <- matrix(-1, length(ulin), length(utipstate))
export_tbl_2 <- matrix(-1, length(ulin), length(utipstate))
colnames(import_tbl_1) <- utipstate
colnames(import_tbl_2) <- utipstate
colnames(export_tbl_1) <- utipstate
colnames(export_tbl_2) <- utipstate
rownames(import_tbl_1) <- ulin
rownames(import_tbl_2) <- ulin
rownames(export_tbl_1) <- ulin
rownames(export_tbl_2) <- ulin
for (i in 1:length(ulin)) {
  if ((sub_stats[i,1] <= 4000) & ((sub_stats[i,2]>0) | (sub_stats[i,3]>0))) {
    sub_tree <- sts[[i]]
    sub_tree <- multi2di(sub_tree)
    einds    <- which(sub_tree$edge.length <=  etol)
    sub_tree$edge.length[einds] <- etol
    sub_tree <- multi2di(sub_tree)
    
    tipstate <- paste(sub_tree$continent)
    tipstate[which(sub_tree$country=="India")] <- "India"
    tipstate[which(sub_tree$country=="China")] <- "China"
    tipstate[which(sub_tree$country=="Italy")] <- "Italy"
    tipstate[which(sub_tree$country=="Spain")] <- "Spain"
    tipstate[which(sub_tree$country=="Singapore")] <- "Singapore"
    tipstate[which(sub_tree$country=="UK")]    <- paste(sub_tree$country_state[which(sub_tree$country=="UK")])
    ace_ER <- ace(tipstate, sub_tree, type="discrete", model="ER")
    sub_tree$tipstate <- tipstate
    anc_state <- colnames(ace_ER$lik.anc)[apply(ace_ER$lik.anc, 1, which.max)]
    sub_tree$state  <- factor(c(tipstate, anc_state),levels=utipstate)
    sub_tree$ace_ER <- ace_ER
    sts[[i]] <- sub_tree
    
    #print(ulin[i])
    #print( table(sub_tree$state[sub_tree$edge[,1]], sub_tree$state[sub_tree$edge[,2]]) )
    
    for (j in 1:length(focus_states)) {
      from_inds <- which(sub_tree$state[sub_tree$edge[,1]]==focus_states[j])
      if (length(from_inds)>0) {
        exports_from_focus <- table( sub_tree$state[sub_tree$edge[from_inds,2]]  )
      } else {
        exports_from_focus <- array(0, length(utipstate))
      }
      
      to_inds   <- which(sub_tree$state[sub_tree$edge[,2]]==focus_states[j])
      if (length(to_inds)>0) {
        imports_to_focus <- table( sub_tree$state[sub_tree$edge[to_inds,1]]  )
      } else {
        imports_to_focus <- array(0, length(utipstate))
      }
      
      if (j==1) {
        export_tbl_1[i,] <- exports_from_focus
        import_tbl_1[i,] <- imports_to_focus
      } else {
        export_tbl_2[i,] <- exports_from_focus
        import_tbl_2[i,] <- imports_to_focus
      }
    }
    
  }
}
```

```{r, echo=FALSE}
kable(sub_stats, caption="Number of sequences in each subtree")
```

```{r, echo=FALSE}
kable(import_tbl_1, caption="Imports to Scotland")
```

```{r, echo=FALSE}
kable(export_tbl_1, caption="Exports from Scotland")
```

```{r, echo=FALSE}
kable(import_tbl_2, caption="Imports to India")
```

```{r, echo=FALSE}
kable(import_tbl_2, caption="Exports from India")
```
